Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-19 optim.qmd
small tweaks
Update optim.qmd

---
title: Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit
header-includes:
   - \usepackage{algorithm}
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - projection pursuit
  - optimization
  - jellyfish optimiser
  - data visualisation
  - high-dimensional data
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
editor: 
  markdown: 
    wrap: 72
---

@@

The artificial jellyfish search (JS) algorithm [@chou_novel_2021] is a
swarm-based metaheuristic optimisation algorithm inspired by the search
behaviour of jellyfish in the ocean. It is one of the newest swarm
intelligence algorithms [@rajwar_exhaustive_2023], which was shown to
have stronger search ability and faster convergence with few algorithmic
parameters compared to classic optimization methods
[@chou_novel_2021]-[@chou_recent_2022].

Effective optimisation is an important aspect of many methods employed
for visualising high-dimensional data ($X$). Here we are concerned about
computing informative linear projections of high-dimensional ($p$) data
using projection pursuit (PP) (@kr69, @FT74). This involves optimising a
function (e.g. @hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020), called the
projection pursuit index (PPI), that defines what is interesting or
informative in a projection.

These PPI are defined on projections ($XA$), which means that there is a
constraint that needs to be considered when optimising. A projection of
data is defined by a $p\times d$ orthonormal matrix $A$, and this
imposes the constraint on the elements of $A$, that columns need have
norm equal to 1 and the product of columns need to sum to zero.

@cook1995grand introduced the PP guided tour, which enabled interactive
visualisation of the optimisation in order to visually explore
high-dimensional data. It is implemented in the R [@R] package `tourr`
[@tourr]. The optimisation that is implemented is fairly basic, and
potential problems were highlighted by @RJ-2021-105. Implementing better
optimisation functionality is a goal, but it needs to be kept in mind
that the guided tour also has places importance on watching the
projected data as the optimisation progresses.

Here we explore the potential for a jellyfish optimisation to be
integrated with the guided tour. @sec-background explains the
optimisation that is used in the current the projection pursuit guided
tour. @sec-theory provides more details on the jellyfish optimiser and
formalises several characteristics of projection pursuit indexes that
are help to measure optimiser performance. @sec-simulation describes a
simulation study on performance of the jellyfish for several types of
data and index functions. @sec-conclusion summarises the work and
provides suggestions for future directions.

@@

```{r}
dplyr::glimpse(sim_sine_6d_spline_head |> mutate(index = "spline2d") |> select(-id, -info)) 
```

>>>>

---
title: Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit
header-includes:
   - \usepackage{algorithm}
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - projection pursuit
  - optimization
  - jellyfish optimiser
  - data visualisation
  - high-dimensional data
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
editor: 
  markdown: 
    wrap: 72
---

@@

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely *projection
pursuit index (PPI)*, (@hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020). Given
high-dimensional data $X \in \mathcal{R}^{n }\times p$ and PPI
\$f(\\cdot)\$, PP finds the orthonormal projection matrix
$A \in \mathcal{R}^{p \times d}$ by solving the following optimisation
problem:

$$
\underset{A}{\max} f(XA) \quad \text{subject to} \quad A'A = I
$$

These indices $f(\cdot)$ are characterised by the informativeness or
“interestingness” of a projection and are often non-linear and
non-convex. As such, an effective and efficient optimisation procedure
is essential to enable sufficient exploration of the data landscape
during the visualisation process and to arrive at a global optimal
viewpoint of the data.

Significant efforts have been devoted to enhancing the functionality of
PP and facilitating a better experience in the visualisation process.
@cook1995grand introduced the PP guided tour, which enabled interactive
visualisation of the optimisation to visually explore high-dimensional
data. It is implemented in the R [@R] package `tourr`[@tourr].
@RJ-2021-105 highlighted potential problems of the optimisation routine
implemented. While improving the quality of the optimisation solutions
in the tour is essential, it is also important to be able to watch the
projected data as the optimisation progresses. Therefore, it is of
interest to integrate the guided tour with a global optimisation
algorithm and examine the characteristics and advantages of such
implementation.

The artificial jellyfish search (JS) algorithm [@chou_novel_2021] is a
swarm-based metaheuristic optimisation algorithm inspired by the search
behaviour of jellyfish in the ocean. It is one of the latest swarm
intelligence algorithms [@rajwar_exhaustive_2023], which was shown to
have stronger search ability and faster convergence with fewer
parameters compared to classic optimisation methods
[@chou_novel_2021]-[@chou_recent_2022].

This study aims to investigate ...

The contribution of this study to the high dimensional data visualisation literature is twofold. First, the performance of JSO is studied in the context of PP guided tour.
Second, to the best of the authors’ knowledge, this is the first attempt
to quantitatively measure squintability and smoothness.

The rest of this paper is structured as follows. @sec-background
introduces the PP guided tour. @sec-theory describes the details of JSO
and introduces metrics that measure different characteristics of a PPI.
@sec-simulation describes a simulation study on the performance of the
JSO using well-known data sets and index functions. @sec-conclusion
summarises the work and provides suggestions for future directions.

@@

```{r}
dplyr::glimpse(sim_sine_6d_spline_head |> select(-id, -info)) 
```
